Tesseroids: forward modeling gravitational fields in spherical coordinates

### 2015-03-13 manuscript.tex
More comments and corrections from @valcris

\address{
\footnotemark[1]Universidade do Estado do Rio de Janeiro,
Rio de Janeiro, Brazil.
email: leouieda@gmail.com
\\
\footnotemark[2]Observat\'orio Nacional,
Rio de Janeiro, Brazil.
\\
\footnotemark[3]Department of Mathematics and Geosciences, University of
Trieste, Trieste, Italy
}

@@

Figure~\ref{fig:glqerrorsample} shows that the largest errors are spread over
on top of the tesseroid.
Thus, calculating the tesseroid fields at a single point might not
capture the point of largest error.
Instead, we calculate the effect of the tesseroid model on a regular grid
of $10 \times 10$ points at different geographic locations
(see Table~\ref{tbl:experiment}).
\todo[inline]{ VAL: A grid of 10 X  10  point covers a are of ??? X ???? km}
\todo[inline]{LEO: The grid location varied in each experiment of the numerical analysis. Table 1 shows the locations of the grids.
This is discussed below.
I'll add this information to the text.}
Fortunately, the symmetry of the shell allows us to consider the computation
point at any geocentric coordinate.
Therefore, the effect of the shell will be same along the entire grid.
We compute the differences between the effects of the shell and the tesseroid
model on the grid.
However, we will consider only the largest error in our analysis.
\todo[inline]{ VAL: I rewrite the above sentence by:
"We compute the differences between the effects of the shell and the tesseroid model on the grid; however, we consider  the largest error only.  }
\todo[inline]{LEO: Rewritten. Is it better now?}
We placed the grid on top of a particular tesseroid
to increase the chances of capturing the true largest integration error.
\todo[inline]{ VAL:I rewrite the above sentence as:
"We decided for the nonconservative procedure of placing the grid on top of a particular tesseroid to increase the chances
of capturing the true largest integration error"}
\todo[inline]{LEO: Changed. I don't think this is really
``nonconservative'' so I left it out.}

\todo[inline]{ I've stopped here - VALERIA}

We calculate the errors for values of the distance-size ratio
$D$ varying from 0 (i.e., no divisions) to 10 in 0.5 intervals.
Furthermore, we repeated the analysis in four different numerical experiments,
each with computation grids at different locations and
different tesseroid model sizes.
Table~\ref{tbl:experiment} describes the different numerical
experiments and the corresponding parameters of the computation grid and
tesseroid model.

@@

For the potential $V$, a distance-size ratio $D=1$ guarantees that the curves
for all experiments are below the $0.1\%$ error threshold.
For $g_z$, the same is achieved with $D=1.5$.
Conversely, $g_{zz}$ requires a value of $D=8$ to achieve an error level of
$0.1\%$.
For a computation height of 260 km, the error curve for $G_{zz}$
intercepts the error threshold line at $D=2.5$.
This behavior suggests that the error curves for $g_{zz}$ might depend on the
computation height.
To test this hypothesis, we computed the error curves for $g_{zz}$ at
heights 2, 10, 50, 150, and 260 km.
Figure~\ref{fig:gzz-with-height} shows the results for $g_{zz}$ at varying
computation heights.
Notice that the distance-size ratio needed to achieve $0.1\%$ accuracy
decreases as the computation height increases.
For example, computation at 260 km height requires $D=2.5$ whereas at 10 km
height a value of $D=5.5$ is required to achieve the same accuracy.
One can take advantage of this behavior to reduce the distance-size ratio for
computations of the gravity gradient tensor at high altitudes,
saving computation time.

@@

\begin{table*}
\centering
\begin{tabular}{cccc}
    \hline
    ~             & Grid location & Grid height & Tesseroid size \\
    \hline
    Experiment 1  & 89N--90N/0E--1E & 2 km & $1^\circ \times 1^\circ$ \\
    Experiment 2  & 0N--1N/0E--1E & 2 km & $1^\circ \times 1^\circ$ \\
    Experiment 3  & 89N--90N/0E--1E & 260 km & $1^\circ \times 1^\circ$ \\
    Experiment 4  & 60N--90N/0E--30E & 2 km & $30^\circ \times 30^\circ$ \\
    \hline
\end{tabular}
\caption{
    Parameters of the numerical experiments to quantify the accuracy of
    the numerical integration.
    All grids had $10\times10$ regularly spaced computation points at a
    constant height.
    Tesseroids used to discretize the spherical shell had 1 km thickness and
    the horizontal dimensions shown in the table.
}
\label{tbl:experiment}
\end{table*}

@@

@@

We performed an error analysis to determine
the optimal value of $D$ required to achieve a target accuracy.
We used a spherical shell as a reference to calculate the computation error of
our algorithm for different values of $D$.
Our results show that the values of $D$ required to achieve a maximum error
of 0.1\% of the shell values are
1 for the gravitational potential, 1.5 for the gravitational acceleration,
and 8 for the gravity gradients.
Previous assumptions were that accurate results are guaranteed if
the distance to the tesseroid is larger than
the distance between point masses.
This same condition was previously applied indiscriminately
to both the gravitational acceleration and the gravity gradients.
That assumption is equivalent to using $D=1.5$ for all fields.
Our results show that this is valid for the gravitational
acceleration and results in a 0.1\% computation error.
This is expected because the original study that determined the above condition
was performed on the vertical component of gravitational acceleration.
However, applying the same condition to the gravity gradients produces
an error of the order of $10^2\%$.

@@

The authors were supported in this research by
a fellowship (VCFB) from
Conselho Nacional de Desenvolvimento Cient\'ifico e Tecnol\'ogico (CNPq)
and a scholarship (LU) from
Coordena\c{c}\~ao de Aperfei\c{c}oamento de Pessoal de N\'ivel Superior
(CAPES),
Brazil.
Additional support for the authors was provided by
the Brazilian agencies CNPq (grant 471693/2011-1) and
FAPERJ (grant E-26/103.175/2011).

>>>>

\address{
\footnotemark[1]Universidade do Estado do Rio de Janeiro,
Rio de Janeiro, Brazil.
email: leouieda@gmail.com
\\
\footnotemark[2]Observat\'orio Nacional,
Rio de Janeiro, Brazil.
email: valcris@on.br
\\
\footnotemark[3]Department of Mathematics and Geosciences, University of
Trieste, Trieste, Italy
}

@@

Figure~\ref{fig:glqerrorsample} shows that the largest errors are spread over
on top of the tesseroid.
Thus, calculating the tesseroid fields at a single point might not
capture the point of largest error.
Instead, we calculate the effect of the tesseroid model on a regular grid
of $10 \times 10$ points at different geographic locations
(see Table~\ref{tbl:experiment}).
\todo[inline]{ VAL: A grid of 10 X  10  point covers a are of ??? X ???? km}
\todo[inline]{LEO: The grid location varied in each experiment of the numerical analysis. Table 1 shows the locations of the grids.
This is discussed below.
I'll add this information to the text.}
Fortunately, the symmetry of the shell allows us to consider the computation
point at any geocentric coordinate.
Therefore, the effect of the shell will be same along the entire grid.
We compute the differences between the effects of the shell and the tesseroid
model on the grid.
However, we will consider only the largest error in our analysis.

\todo[inline]{ VAL: I rewrite the above sentence by:
"We compute the differences between the effects of the shell and the tesseroid model on the grid; however, we consider  the largest error only.  }
\todo[inline]{LEO: Rewritten. Is it better now?}
We placed the grid on top of a particular tesseroid
to increase the chances of capturing the true largest integration error.
\todo[inline]{ VAL:I rewrite the above sentence as:
"We decided for the nonconservative procedure of placing the grid on top of a particular tesseroid to increase the chances
of capturing the true largest integration error"}
\todo[inline]{LEO: Changed. I don't think this is really
``nonconservative'' so I left it out.}

We calculate the errors for values of the distance-size ratio
$D$ varying from 0 (i.e., no divisions) to 10 in 0.5 intervals.
Furthermore, we repeated the error analysis in four different numerical experiments,
each with computation grids at different locations and
different tesseroid model sizes.
Table~\ref{tbl:experiment} describes the different numerical
experiments and the corresponding parameters of the computation grid and
tesseroid model.

 \todo[inline]{ Val: Regarding the visual aspect of Figure 4. I think that  the dashed lines are yielding a visual pollution. However, they are helping us to identify the values in the axes. Regarding the visual aspect of Figure 4.
A possibility would be keep the dashed lines, but with  other color (e.g., light gray  dashed lines) }

@@

For the potential $V$, a distance-size ratio $D=1$ guarantees that the curves
for all experiments are below the $0.1\%$ error threshold.
For $g_z$, the same is achieved with $D=1.5$.
Conversely, $g_{zz}$ requires a value of $D=8$ to achieve an error level of
$0.1\%$.
For a computation height of 260 km, the error curve for $g_{zz}$
intercepts the error threshold line at $D=2.5$.
This behavior suggests that the error curves for $g_{zz}$ might depend on the
computation height.
To test this hypothesis, we computed the error curves for $g_{zz}$ at
heights 2, 10, 50, 150, and 260 km.
Figure~\ref{fig:gzz-with-height} shows the results for $g_{zz}$ at varying
computation heights.
Notice that the distance-size ratio required to achieve $0.1\%$ accuracy
decreases as the computation height increases.
For example, computation at 260 km height requires $D=2.5$ whereas at 10 km
height a value of $D=5.5$ is required to achieve the same accuracy.
One can take advantage of this behavior to reduce the distance-size ratio for
computations of the gravity gradient tensor at high altitudes,
saving computation time.

@@

\todo[inline]{ VAL: I think that we discussed that similar results that are achieved in gzz (Figure 5) also are I achieved in gxx, gyy.... I think that this information could be written.}

\begin{table*}
\centering
\begin{tabular}{lccc}
    \hline
    ~             & Grid location & Grid height & Tesseroid size \\
    \hline
    Experiment 1 (pole)            & 89N--90N/0E--1E  & 2 km   & $1^\circ \times 1^\circ$ \\
    Experiment 2 (equator)         & 0N--1N/0E--1E    & 2 km   & $1^\circ \times 1^\circ$ \\
    Experiment 3 (260 km)          & 89N--90N/0E--1E  & 260 km & $1^\circ \times 1^\circ$ \\
    Experiment 4 ($30^\circ$ size) & 60N--90N/0E--30E & 2 km   & $30^\circ \times 30^\circ$ \\
    \hline
\end{tabular}
\caption{
    Parameters of the numerical experiments to quantify the accuracy of
    the numerical integration.
    All grids had $10\times10$ regularly spaced computation points at a
    constant height.
    Tesseroids used to discretize the spherical shell had 1 km thickness and
    the horizontal dimensions shown in the table.
}
\label{tbl:experiment}
\end{table*}

@@

\todo[inline]{ I am here. I did not read the captions of the Figures 4 and 5 and the caption of the Table 1}

@@

We performed an error analysis to determine
the optimal value of $D$ required to achieve a target accuracy.
We used a spherical shell as a reference to calculate the computation error of
our algorithm for different values of $D$.
Our results show that the values of $D$ required to achieve a maximum error
of 0.1\% of the shell values are
1 for the gravitational potential, 1.5 for the gravitational acceleration,
and 8 for the gravity gradients.
Previous assumptions in the literature
were that accurate results are guaranteed if
the distance to the tesseroid is larger than
the distance between point masses.
This condition was previously applied indiscriminately
to both the gravitational acceleration and the gravity gradients.
That assumption is equivalent to using $D=1.5$ for all fields.
Our results show that this is valid for the gravitational
acceleration and results in a 0.1\% computation error.
This is expected because the original study that determined the above condition
was performed on the vertical component of gravitational acceleration.
However, applying the same condition to the gravity gradients produces
an error of the order of $10^2\%$.

@@

The authors were supported in this research by
a fellowship (VCFB) from
Conselho Nacional de Desenvolvimento Cient\'ifico e Tecnol\'ogico (CNPq)
and a scholarship (LU) from
Coordena\c{c}\~ao de Aperfei\c{c}oamento de Pessoal de N\'ivel Superior
(CAPES),
Brazil.
Additional support for the authors was provided by
the Brazilian agency FAPERJ (grant E-26/103.175/2011).
